---
title: "Exploring US Births"
output:
  html_notebook:
    toc: true
    toc_float: true
  html_document:
    toc: true
    toc_depth: 2
    theme: united
    highlight: tango
  github_document:
    html_preview: false
    toc: true
    toc_depth: 2
---

```{r global_options, include=FALSE}

knitr::opts_chunk$get()
knitr::opts_chunk$set(echo=TRUE, warning=FALSE, message=FALSE)

rm(list = ls())

### run get_data.R first if data/ directory not found
if (!file.exists("data")) { 
     source("get_data.R")
}
```


## Summary

```{r}
file_ssa <- "./data/US_births_2000-2014_SSA.csv"
file_cdc <- "./data/US_births_1994-2003_CDC_NCHS.csv"
births2000_2014 <- read.csv(file_ssa)
births1994_2003 <- read.csv(file_cdc)
```

```{r}
str(births1994_2003)
summary(births1994_2003)
```

```{r}
str(births2000_2014)
summary(births2000_2014)
```


## Prepare the data

First of all, we should visually check to see if the number of babies born match for dates between 2000 and 2003.
```{r}
head(subset(births1994_2003, year = 2003), n=3); head(subset(births2000_2014, year = 2003), n=3)
head(subset(births1994_2003, year = 2002), n=3); head(subset(births2000_2014, year = 2002), n=3)
head(subset(births1994_2003, year = 2000), n=3); head(subset(births2000_2014, year = 2000), n=3)
```

Clearly the birth numbers vary. We'll merge the dataframes using only the SSA data where the dates intersect (2000 to 2003).

```{r}
births_all <- rbind(births2000_2014, subset(births1994_2003, year < 2000))
```

```{r, echo=FALSE, results='hide'}
sort(unique(births_all$month))
sort(unique(births_all$year))
sort(unique(births_all$date_of_month))
sort(unique(births_all$day_of_week))
```


## In general, what month(s) encourage more pregnancies? 

Investigate patterns related to when (months, seasons) people decide to get pregnant.

It would be nice to find out if most people generally get pregnant in the late fall early winter months as that is the general belief. To keep things simple, we'll assume that all pregnancies last exactly 40 weeks.

```{r}
# load libraries
source("packages.R")

year_str <- as.character(births_all$year)
month_str <- as.character(births_all$month)
date_str <- as.character(births_all$date_of_month)
full_date_str <- paste0(year_str, '/', month_str, '/', date_str)
births_all$pregnancy_start <- ymd(full_date_str) - weeks(40)

plot_colours <- brewer.pal(8, "Set2")   # for bar plots
boxplot_colours <- brewer.pal(9, "YlGn")   # for monthly boxplots
other_colours <- brewer.pal(4, "Dark2")   #
```

```{r}
### yearly total births
yearly_births <- births_all %>% group_by(year) %>% summarize(total_births = sum(births))

ggplot(yearly_births, aes(year, total_births)) +
     geom_bar(stat = "identity", fill=plot_colours[3]) + 
     labs(y="births", x="", title="Yearly total US births") +
     theme(plot.title = element_text(hjust = 0.5))   # center the tile
```

```{r}
births_all$month_lab <- month(births_all$pregnancy_start, label = TRUE)

### Total pregnancies - by month, from 1994 to 2014.
monthly_births <- births_all %>% group_by(month_lab) %>% summarize(total_births = sum(births))

ggplot(monthly_births, aes(month_lab, total_births)) +
     geom_bar(stat = "identity", fill=plot_colours[1]) + 
     geom_hline(aes(yintercept=median(total_births)), colour=other_colours[2]) +   # median line
     labs(y="pregnancies", x="", title="Monthly total US pregnancies") + 
     theme(plot.title = element_text(hjust = 0.5))   # center the tile

### Average number of monthly pregnancies
avg_monthly_births <- births_all %>% group_by(month_lab) %>% summarize(avg_births = mean(births))
avg_monthly_median <- median(avg_monthly_births$avg_births)

ggplot(avg_monthly_births, aes(month_lab, avg_births)) +
     geom_bar(stat = "identity", fill=plot_colours[2]) + 
     geom_hline(aes(yintercept=avg_monthly_median), colour=other_colours[1]) +   # median line
     labs(y="pregnancies", x="", title="Monthly average US pregnancies") + 
     theme(plot.title = element_text(hjust = 0.5))   # center the tile
```

```{r}
ggplot(births_all, aes(month_lab, births, fill=month_lab)) +
     geom_boxplot() +
     guides(fill=FALSE) +
     labs(y="pregnancies", x="", title="Monthly US pregnancies") +
     theme(plot.title = element_text(hjust = 0.5))   # center the tile
```


There is a gradual increase in the number of pregnanices as we go from summer to autumn/fall. Looking at the average monthly plot, the fall and winter months (September to January) are the highest above the median line. The average for May is higher than the expected trend, relative to April and June. This could be explained by the fact that some professions, for example teachers, do not have an extended vacation any other time of year.

The monthly boxplots seem to confirm the general trend of increasing pregnancies starting around August and peaking in December.


## Compare number of births from the two sets of data for the years 2000 to 2003 - yearly, monthly

This will highlight the general pattern of the birth numbers from both data sets
```{r}
df1 <- subset(births2000_2014, year < 2004)
names(df1) <- paste0(names(df1), "_1")
df2 <- subset(births1994_2003, year > 1999)
names(df2) <- paste0(names(df2), "_2")

births_intersection <- cbind(df1, df2)   
```

```{r, echo=FALSE, results='hide'}
sort(unique(births_intersection$year_1))
sort(unique(births_intersection$year_2))

# check to make sure the dates for the numbers check out
which(births_intersection$year_1 != births_intersection$year_2 | 
           births_intersection$month_1 != births_intersection$month_2 | 
           births_intersection$date_of_month_1 != births_intersection$date_of_month_2)
```

```{r}
# merge columns, cleanup
year_eq <- all(births_intersection$year_1 == births_intersection$year_2)
month_eq <- all(births_intersection$month_1 == births_intersection$month_2)
month_date_eq <- all(births_intersection$date_of_month_1 == births_intersection$date_of_month_2)
if (year_eq) {
     births_intersection$year_2 <- NULL
}
if (month_eq) {
     births_intersection$month_2 <- NULL
}
if (month_date_eq) {
     births_intersection$date_of_month_2 <- NULL
     births_intersection$day_of_week_2 <- NULL
}

rm(year_eq, month_eq, month_date_eq, df1, df2)

# rename columns, reshape data, cleanup
names(births_intersection)[1:4] <- sub("_1", '', names(births_intersection)[1:4])
names(births_intersection)[5] <- "SSA"
names(births_intersection)[6] <- "CDC"

births_intersection <- melt(births_intersection, id.vars = 1:4, 
                            variable.name = "source", value.name = "births")
head(births_intersection)
tail(births_intersection)

year_str <- as.character(births_intersection$year)
month_str <- as.character(births_intersection$month)
date_str <- as.character(births_intersection$date_of_month)
full_date_str <- paste0(year_str, '/', month_str, '/', date_str)
births_intersection$full_date <- ymd(full_date_str)
births_intersection$month_lab <- month(births_intersection$full_date, label = TRUE)

rm(year_str, month_str, date_str, full_date_str)
```


```{r}
### compare yearly total births
yearly_births_2 <- births_intersection %>% group_by(year, source) %>% summarize(total_births = sum(births))
ggplot(yearly_births_2, aes(year, total_births, fill=source)) +
     geom_bar(stat = "identity", position = position_dodge()) + 
     scale_fill_manual(values=plot_colours[c(3,8)]) +
     labs(y="births", x="", title="Yearly total US births") +
     theme(plot.title = element_text(hjust = 0.5))   # center the tile

### compare avearge monthly total births
avg_monthly_births_2 <- births_intersection %>% group_by(month_lab, source) %>%
     summarize(avg_births = mean(births))
ggplot(avg_monthly_births_2, aes(month_lab, avg_births, fill=source)) +
     geom_bar(stat = "identity", position = position_dodge()) + 
     scale_fill_manual(values=plot_colours[c(2,8)]) +
     geom_hline(aes(yintercept=avg_monthly_median), colour=other_colours[4]) +   # median line
     labs(y="births", x="", title="Monthly average US births") +
     theme(plot.title = element_text(hjust = 0.5))   # center the tile
```

```{r}
ggplot(births_intersection, aes(source, births, fill=source)) +
     geom_boxplot() + coord_flip() +
     guides(fill=FALSE) +
     scale_fill_manual(values = other_colours[3:4]) +
     labs(y="births", x="", title="US births by data source") +
     theme(plot.title = element_text(hjust = 0.5))   # center the tile
```

It looks like the birth numbers from the SSA are consistently higher than those from the CDC. This might be worth further investigation.

## Conclusion
There is noticeable increase in pregnancies that start in the fall or early winter months. A lot of the this increase is witnessed in the months of October, November, and December.  

